Place of death in NRS COVID-19 deaths: data processing for analysis

### 2022-01-13 covid_nrs_place_of_death_data.Rmd
added table output for SPPC data request - summary of 2020 and 2021 data

## bespoke computation because format is different... deaths split by cause!
total_deaths_cause_of_death <-
  merged_deaths_cause_of_death %>%
  group_by(year, place_of_death, cause_of_death) %>%
  summarise(deaths = sum(number_of_deaths), .groups="drop") %>%
  left_join(historical_annual_average_cause_of_death, by = c("place_of_death", "cause_of_death")) %>%
  mutate(
    ratio_annual_deaths_to_historical_mean = deaths / annual_deaths_mean
  )
```

@@

```{r}
table_current_deaths_compared_to_past <-
  bind_rows(
    ## all available data
    merged_deaths_overall %>%
      filter(age == "all" &
               sex == "all"
             ) %>% 
      group_by(year, place_of_death, week_number) %>%
      summarise(
        deaths_all_causes = sum(deaths_all_causes),
        # deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        .groups = "drop"
      ) %>% 
      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
      left_join(
        past_deaths_overall %>% filter(year != "average") %>% group_by(week_number, place_of_death) %>% summarise(
                                           average_weekly_deaths = sum(number_of_deaths) / 5,
                                           .groups = "drop"
                                         ),
        by = c("week_number", "place_of_death")
      ) %>% 
      group_by(place_of_death) %>%
      summarise(
        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
        deaths_all_causes = sum(deaths_all_causes),
        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
        .groups = "drop"
      )
    ,
    ## 2020 only
    merged_deaths_overall %>%
      filter(age == "all" &
               sex == "all" &
               year == 2020
             ) %>% 
      group_by(place_of_death) %>%
      summarise(
        period = "2020",
        deaths_all_causes = sum(deaths_all_causes),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        .groups = "drop"
      ) %>% 
      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
    left_join(
      historical_annual_average_overall %>% 
        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
      by = "place_of_death"
    )
    ,
    ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
    merged_deaths_overall %>%
      filter(
        age == "all" &
          sex == "all" &
          year == 2020 & week_number >= 12
      ) %>%
      group_by(place_of_death) %>%
      summarise(
        period = "2020, week 12 - week 53",
        deaths_all_causes = sum(deaths_all_causes),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        .groups = "drop"
      ) %>% 
    mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
    left_join(
      past_deaths_overall %>% 
        filter(week_number >= 12 & year != "average") %>%
        group_by(place_of_death) %>%
        summarise(mean_annual_deaths_2015_2019 = sum(number_of_deaths) / 5),
      by = "place_of_death"
    )
  ) %>%
  mutate(
    ratio_deaths_all_causes_to_historical = deaths_all_causes / mean_annual_deaths_2015_2019,
    ratio_non_covid_deaths_to_historical = deaths_non_covid / mean_annual_deaths_2015_2019,
    proportion_covid_deaths = deaths_covid_related / deaths_all_causes
  )

@@

@@

```{r}
total_deaths_overall %>%
  mutate(excess_deaths = deaths_all_causes - annual_deaths_min)

excess_deaths_2020_table <-
  past_deaths_overall %>%
  group_by(year, place_of_death) %>%
  summarise(annual_deaths = sum(number_of_deaths), .groups = "drop") %>%
  filter(year==2019) %>%
  pivot_wider(names_from = year, names_prefix = "annual_deaths_", values_from = annual_deaths) %>%
  left_join(
    total_deaths_overall %>% filter(year==2020), by = "place_of_death"
    ) %>%
  mutate(
    excess_deaths_vs_2019 = deaths_all_causes - annual_deaths_2019,
    # excess_deaths_vs_2019_prop = excess_deaths_vs_2019 / sum(excess_deaths_vs_2019),
    excess_deaths_vs_mean = deaths_all_causes - annual_deaths_mean,
    excess_deaths_vs_min = deaths_all_causes - annual_deaths_min,
    excess_deaths_vs_max = deaths_all_causes - annual_deaths_max
  ) %>%
  select(year, place_of_death, annual_deaths_2019, annual_deaths_mean, deaths_all_causes, matches("excess")) %>%
  mutate(disaggregated_place_of_death = place_of_death != "All") %>%
  group_by(disaggregated_place_of_death) %>%
  mutate(
    proportion_excess_deaths_vs_2019 = excess_deaths_vs_2019 / sum(excess_deaths_vs_2019),
    proportion_excess_deaths_vs_mean = excess_deaths_vs_mean / sum(excess_deaths_vs_mean)
  ) %>%
  ungroup %>%
  select(-disaggregated_place_of_death)

>>>>

## bespoke computation because format is different... deaths split by cause!
total_deaths_cause_of_death <-
  merged_deaths_cause_of_death %>%
  group_by(year, place_of_death, cause_of_death) %>%
  summarise(deaths = sum(number_of_deaths), .groups="drop") %>%
  left_join(historical_annual_average_cause_of_death, by = c("place_of_death", "cause_of_death")) %>%
  mutate(
    ratio_annual_deaths_to_historical_mean = deaths / annual_deaths_mean
  )

current_and_historical_deaths_by_cause_and_location <-
  bind_rows(
    ## 2020 & 2021 separately
    total_deaths_cause_of_death %>%
      select(year, place_of_death, cause_of_death, deaths) %>%
      mutate(year = as.character(year)),
    ## 2020 & 2021 averaged
    total_deaths_cause_of_death %>%
      select(year, place_of_death, cause_of_death, deaths) %>%
      group_by(place_of_death,  cause_of_death) %>%
      summarise(deaths = sum(deaths), year = "2020-2021", .groups = "drop"),
    ## historical data
    total_deaths_cause_of_death %>%
      select(year, place_of_death, cause_of_death, deaths = annual_deaths_mean) %>%
      mutate(year = "2015-2019") %>%
      distinct
  ) %>%
  filter(place_of_death != "All") %>%
  group_by(year, cause_of_death) %>%
  mutate(proportion = deaths / sum(deaths)) %>%
  ungroup %>%
  arrange(year, cause_of_death)

```

@@

```{r}
table_current_deaths_compared_to_past <-
  bind_rows(
    ## all available data
    merged_deaths_overall %>%
      filter(age == "all" &
               sex == "all"
             ) %>% 
      group_by(year, place_of_death, week_number) %>%
      summarise(
        deaths_all_causes = sum(deaths_all_causes),
        # deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        .groups = "drop"
      ) %>% 
      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
      left_join(
        past_deaths_overall %>% filter(year != "average") %>% group_by(week_number, place_of_death) %>% summarise(
                                           average_weekly_deaths = sum(number_of_deaths) / 5,
                                           .groups = "drop"
                                         ),
        by = c("week_number", "place_of_death")
      ) %>% 
      group_by(place_of_death) %>%
      summarise(
        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
        deaths_all_causes = sum(deaths_all_causes),
        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
        .groups = "drop"
      )
    ,
    ## 2020 only
    merged_deaths_overall %>%
      filter(age == "all" &
               sex == "all" &
               year == 2020
             ) %>% 
      group_by(place_of_death) %>%
      summarise(
        period = "2020",
        deaths_all_causes = sum(deaths_all_causes),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        .groups = "drop"
      ) %>% 
      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
    left_join(
      historical_annual_average_overall %>% 
        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
      by = "place_of_death"
    )
    ,
    ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
    merged_deaths_overall %>%
      filter(
        age == "all" &
          sex == "all" &
          year == 2020 & week_number >= 12
      ) %>%
      group_by(place_of_death) %>%
      summarise(
        period = "2020, week 12 - week 53",
        deaths_all_causes = sum(deaths_all_causes),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        .groups = "drop"
      ) %>% 
    mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
    left_join(
      past_deaths_overall %>% 
        filter(week_number >= 12 & year != "average") %>%
        group_by(place_of_death) %>%
        summarise(mean_annual_deaths_2015_2019 = sum(number_of_deaths) / 5),
      by = "place_of_death"
    )
    ,
    ## 2021 only
    merged_deaths_overall %>%
      filter(age == "all" &
               sex == "all" &
               year == 2021
             ) %>% 
      group_by(place_of_death) %>%
      summarise(
        period = "2021",
        deaths_all_causes = sum(deaths_all_causes),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        .groups = "drop"
      ) %>% 
      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
    left_join(
      historical_annual_average_overall %>% 
        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
      by = "place_of_death"
    )
  ) %>%
  mutate(
    ratio_deaths_all_causes_to_historical = deaths_all_causes / mean_annual_deaths_2015_2019,
    ratio_non_covid_deaths_to_historical = deaths_non_covid / mean_annual_deaths_2015_2019,
    proportion_covid_deaths = deaths_covid_related / deaths_all_causes
  )

@@

# Yearly death rates

```{r}
table_annual_death_rates <-
  merged_deaths_overall %>%
    filter(age == "all" &
             sex == "all" &
             year %in% c(2020:2021)
           ) %>% 
    group_by(year, place_of_death) %>%
    summarise(
      deaths_all_causes = sum(deaths_all_causes),
      deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
      .groups = "drop"
    ) %>% 
    mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
  left_join(
    historical_annual_average_overall %>% 
      select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
    by = "place_of_death"
  ) %>%
  mutate(
    proportion_covid_deaths = scales::percent(deaths_covid_related / deaths_all_causes, accuracy = 0.1),
    change_deaths_all_causes_from_historical = scales::percent((deaths_all_causes / mean_annual_deaths_2015_2019) - 1, accuracy = 0.1),
    change_non_covid_deaths_from_historical = scales::percent((deaths_non_covid / mean_annual_deaths_2015_2019) - 1, accuracy = 0.1)
  )

table_annual_death_rates_by_cause <-
  total_deaths_cause_of_death %>%
  select(-matches("min|max")) %>%
  filter(year %in% c(2020:2021)) %>%
  mutate(
    change_from_historical_annual_deaths_2015_2019 = scales::percent(ratio_annual_deaths_to_historical_mean - 1, accuracy = 0.1) %>% if_else(. == "Inf", NA_character_, .)
    ) %>%
  mutate(temporary_group_for_all_causes = cause_of_death=="All") %>%
  group_by(year,place_of_death, temporary_group_for_all_causes) %>%
  mutate(
    proportion_annual_deaths_in_place = scales::percent(deaths / sum(deaths), accuracy = 0.1) %>% if_else(. == "100.0%", NA_character_, .),
    historical_proportion_annual_deaths_in_place = scales::percent(annual_deaths_mean / sum(annual_deaths_mean), accuracy = 0.1) %>% if_else(. == "100.0%", NA_character_, .)
    ) %>%
  ungroup %>%
  rename(mean_annual_deaths_2015_2019 = annual_deaths_mean) %>%
  select(-c(ratio_annual_deaths_to_historical_mean, temporary_group_for_all_causes))

notes_annual_death_rates <-
  tibble( notes = c(
    glue::glue("Note: these data are based on weekly figures reported by NRS; years don't align perfectly with calendar years, for example: week 1 of 2020 started on {ISOweek2date('2020-W01-1')}, and week 53 of 2020 ended on {ISOweek2date('2020-W53-7')}; week 1 of 2021 started on {ISOweek2date('2021-W01-1')}, and week 52 of 2021 ended on {ISOweek2date('2021-W52-7')}.")
    )
  )

write.xlsx(
    x = list(
      "2020 + 2021" = table_annual_death_rates,
      "Causes of death (COD)" = table_annual_death_rates_by_cause,
      "notes" = notes_annual_death_rates
      ), 
    file = file.path("./outputs", "annual_death_rates_and_causes_of_death_2020_2021.xlsx")
      )
```

@@

```{r}
excess_deaths_2020_table <-
  past_deaths_overall %>%
  group_by(year, place_of_death) %>%
  summarise(annual_deaths = sum(number_of_deaths), .groups = "drop") %>%
  filter(year==2019) %>%
  pivot_wider(names_from = year, names_prefix = "annual_deaths_", values_from = annual_deaths) %>%
  left_join(
    total_deaths_overall %>% filter(year==2020), by = "place_of_death"
    ) %>%
  mutate(
    excess_deaths_vs_2019 = deaths_all_causes - annual_deaths_2019,
    # excess_deaths_vs_2019_prop = excess_deaths_vs_2019 / sum(excess_deaths_vs_2019),
    excess_deaths_vs_mean = deaths_all_causes - annual_deaths_mean,
    excess_deaths_vs_min = deaths_all_causes - annual_deaths_min,
    excess_deaths_vs_max = deaths_all_causes - annual_deaths_max
  ) %>%
  select(year, place_of_death, annual_deaths_2019, annual_deaths_mean, deaths_all_causes, matches("excess")) %>%
  mutate(disaggregated_place_of_death = place_of_death != "All") %>%
  group_by(disaggregated_place_of_death) %>%
  mutate(
    proportion_excess_deaths_vs_2019 = excess_deaths_vs_2019 / sum(excess_deaths_vs_2019),
    proportion_excess_deaths_vs_mean = excess_deaths_vs_mean / sum(excess_deaths_vs_mean)
  ) %>%
  ungroup %>%
  select(-disaggregated_place_of_death)
